Experience Report: Haskell in Computational Biology

### 2012-03-11 mrfy_experience_report.tex
intro edited

%
% It can't hurt to remind everyone involved exactly what an Experience
% Report is.
%
%
\begin{abstract}
The Experience Report series is
a body of published, refereed, citable \emph{evidence} that functional
programming really works---or descriptions of obstacles that prevent
it from working.   
In~this Experience Report, we show that for computational biologists,
Haskell combines the flexibility and rapid prototyping of a scripting
language with performance comparable to~C++.
The Haskell language and libraries provide powerful abstractions, 
but we couldn't always get them to work!
Despite some speed bumps around profiling, debugging, and the
uncertain status of any given package, we~found Haskell's ecosystem
provided fine support for developing new algorithms in computational
biology.
\end{abstract}

@@

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
For some software, considerations of performance are paramount; this
software is usually written in C~or~C++. 
For other software, considerations of convenience, readability, and
productivity are more important;
this software is usually written in a high-level, dynamically-typed
``scripting'' or domain-specific language like
Perl, Python, Ruby, SPSS, or~R.
In~this paper, we report on an attempt to replace both these families
of languages with Haskell: 
\begin{itemize}
\item
Although had to write a new implementation of a standard
algorithm, the cost was low, and the connection between code and
algorithm is unusually strong (\secref{viterbi}).
The performance of our new implementation is comparable to the
performance of~C++, and parallel performance was easy to achieve
(\secref{perf}). 
\item
As expected, higher-order functions made it unusually easy to
implement and experiment with a family of related algorithms
(\secref{hofs}).
\item
Haskell is surrounded by a penumbra of libraries and tools that
promise powerful abstractions.
Some deliver very high leverage, but others are not ready for prime
time.
We~found it difficult to tell which is which in advance (\secref{penumbra}).
\item
It is possible to use Haskell successfully for research even with a
very modest background in functional programming. 
Modifying our group's C++ code to try out new research ideas has been
difficult;
trying new ideas in Haskell has been easy
(\secref{comparo}).
\item
The only significant impediment that Haskell has presented to our
research has been the difficulty of understanding and improving the
performance of Haskell programs (\secref{awkward-profiling}).
\end{itemize}

@@

Proteins are cellular machinery. They interact with one another and with other 
molecules to carry out the functions of living cells: metabolism, regulation, 
signalling, and more.
A~protein's function is determined by its structure, 
and its structure is determined by the sequence of amino acids that
form the protein.
\remark{Lost: there are only 20-odd amino acids}
The amino-acid sequence is ultimately determined by a sequence of
nucleic acids in DNA, which we call a gene.
Given a genetic sequence, biologists wish to know the cellular
function of the protein that gene codes for.
The best known method of discovering such function is
to find other proteins of 
similar structure, which likely share similar function.
Proteins that share structure and function are expected to be
descended from a common ancestor---in biological terms, \emph{homologous}---and
thus
the problem of identifying proteins similar to a \textit{query sequence} is called 
\textit{homology detection}.

Computational biologists detect homologies by building 
algorithms which, given a \emph{query sequence} of amino acids,
find known proteins of similar structure.
When the similar proteins are formed from amino-acid sequences that
are not too different from the query sequence, the homology can be
detected by
a family of algorithms called 
\textit{hidden Markov models}~\cite{Eddy:1998ut}.
But in real biological systems,
proteins with similar structure and function may be formed from significantly 
different amino-acid sequences, which are not close in edit distance.
Our~research contribution---the MRFy algorithm---can detect homologies
in amino-acid sequences that are only distantly related.
%
%We will explain an 
%algorithm for detecting reasonably similar sequences, and then move on to 
%explain MRFy, our novel approach for detecting more distantly related sequences 
%for proteins that share similar structure and function.
%
To~enable you to understand our experience with Haskell,
we~sketch not only the MRFy algorithm (\secref{mrfy})
but also one of the older hidden-Markov algorithms (\secref{viterbi}),
which is incorporated into~MRFy.

\remark{LOST: We conclude that implementations of mathematically-intensive 
algorithms such as these lend themselves to the Haskell language, particularly 
when the programmer needs a flexible implementation that simplifies 
experimenting with variants of the algorithm.}

% As a brief refresher in the standard dogma of genetics, recall that genes 
% (strands of deoxyribonucleic acid (DNA), which are sequences of the nucleotides 
% adenosine, cytosine, guanine, and thymine, represented by the letters A, C, G, 
% and T) are transcribed into ribonucleic acid (RNA). Some RNA is then translated 
% into the 20 naturally-occuring amino acids, which form peptide chains -- 
% proteins -- which fold into complex structures in the cell. These proteins are 
% cellular machinery, performing the varied functions an organism needs to 
% survive.

% However, determining what newly sequenced genes actually do -- when they become 
% proteins, what structures those proteins fold into, and what functions they 
% perform -- has not kept pace. Determining the atomic coordinates of a newly 
% discovered protein may require as much as two graduate-student-years in a lab! 
% Fortunately, determining the full structure of a protein is not always 
% necessary. Often, we can identify well-known proteins of similar sequence or 
% structure to a new protein, and thus make educated guesses as to the structure 
% and function of the new protein.
% 
% The problem of taking a newly found protein sequence about which only the 
% sequence is known, and determining what existing proteins of known structure 
% and function most closely resemble that new protein, is known as homology 
% detection. When proteins of similar sequence can be found, this problem is 
% largely solved. To solve the homology detection problem, computational 
% biologists have developed fast, approximate methods for determining the 
% structure and function of proteins based on their sequence. The most popular 
% software for homology detection is called HMMER, which uses a hidden Markov 
% model to capture evolutionary change.

% However, when there are no known proteins of similar sequence, only 
% evolutionarily distant proteins that may share similar structure, this problem 
% is called remote homology detection. Hidden Markov model approaches begin to 
% fail as sequence similarity falls off. Recently, Markov random field 
% approaches, which capture non-local interactions in the protein structure, have 
% been shown to perform well. However, these approaches face the challenge of 
% increased computational complexity; the SMURF program exhibits exponential time 
% complexity, while SMURFLite (work by one of the authors) bounds the exponent by 
% simplifying the Markov random field's dependency graph. MRFy is a 
% stochastic search approach to Markov random fields.

@@

%% \subsection{The Viterbi Algorithm}

>>>>

%
% It can't hurt to remind everyone involved exactly what an Experience
% Report is.
%
%
\begin{abstract}
An Experience Report provides {evidence} that functional
programming really works, or it describes obstacles that prevent
functional programming from working.   
Here, we show that for computational biologists,
Haskell combines the flexibility and rapid prototyping of a scripting
language with performance comparable to~C++.
The~Haskell language and libraries provide powerful abstractions, 
but we couldn't always get them to work.
Despite some speed bumps around profiling, debugging, and the
uncertain status of any given package, we~found that Haskell's ecosystem
made it easy for us to develop new algorithms in computational
biology.
\end{abstract}

@@

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
For some software, considerations of performance are paramount; this
software is usually written in C~or~C++. 
For other software, considerations of convenience, readability, and
productivity are more important;
this software is usually written in a high-level
or domain-specific language like
Perl, Python, Ruby, SPSS, or~R.
In~this paper, we report on an attempt to replace both these families
of languages with Haskell: 
\begin{itemize}
\item
Although had to write a new implementation of a standard
algorithm, the cost was low, and the connection between code and
algorithm is unusually strong (\secref{viterbi}).
Our new Haskell code performs competitively with~C++,
and parallel performance was easy to achieve
(\secref{perf}). 
\item
Higher-order functions made it unusually easy to
implement and experiment with a family of related algorithms
(\secref{hofs}).
\item
The only significant impediment that Haskell presented to our
research was the difficulty of understanding and improving the
performance of Haskell programs (\secref{awkward-profiling}).
\item
We~used Haskell successfully even though we are computational
biologists with little background in functional programming.
Modifying our group's C++ code to try new research ideas has been
difficult;
trying new ideas in Haskell has been easy
(\secref{comparo}).
\item
The Haskell language is surrounded by a penumbra of libraries and tools that
promise powerful abstractions.
Some deliver very high leverage, but others are not ready for prime
time.
We~found the two families difficult to distinguish in advance (\secref{penumbra}).
\end{itemize}

@@

Proteins are cellular machinery. They interact with one another and with other 
molecules to carry out the functions of living cells: metabolism, regulation, 
signalling, and so on.
A~protein's function is determined by its structure, 
and its structure is determined by the sequence of amino acids that
form the protein.
The amino-acid sequence is ultimately determined by a sequence of
nucleic acids in DNA, which we call a gene.
Given a genetic sequence, biologists wish to know the cellular
function of the protein that gene codes for.
The best known method of discovering such function is
to find other proteins of 
similar structure, which likely share similar function.
Proteins that share structure and function are expected to be
descended from a common ancestor---in biological terms, \emph{homologous}---and
thus
the problem of identifying proteins similar to a \textit{query sequence} is called 
\textit{homology detection}.

Computational biologists detect homologies by building 
algorithms which, given a {query sequence} of amino acids,
find known proteins of similar structure.
When the similar proteins are formed from amino-acid sequences that
are not too different from the query sequence, the homology can be
detected by
a family of algorithms called 
\textit{hidden Markov models}~\cite{Eddy:1998ut}.
But in real biological systems,
proteins with similar structure and function may be formed from significantly 
different amino-acid sequences, which are not close in edit distance.
Our~research contribution---a software tool called MRFy (pronounced
``Murphy'')---can detect homologies 
in amino-acid sequences that are only distantly related.
%
%We will explain an 
%algorithm for detecting reasonably similar sequences, and then move on to 
%explain MRFy, our novel approach for detecting more distantly related sequences 
%for proteins that share similar structure and function.
%

%
% I hope the new intro renders these two sentences unnecessary. ---NR
%
%
%%  
%%  To~enable you to understand our experience with Haskell,
%%  we~sketch not only the MRFy algorithm (\secref{mrfy})
%%  but also one of the older hidden-Markov algorithms (\secref{viterbi}),
%%  which is incorporated into~MRFy.

@@
